---
title: "Exercise 3: Marketing A/B Testing"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(readr)
```

# 1. Load the data

```{r load-data}
marketing <- read_csv("marketing_AB_mini.csv")

dim(marketing)
head(marketing)
summary(marketing)

table(marketing$`test group`)
table(marketing$converted)
```

The dataset has `r nrow(marketing)` rows and `r ncol(marketing)` columns. The
A/B variable is `test group`, where `ad` is the advertising treatment and `psa`
is the control condition. The main outcome variable is `converted`.

# 2. Research question and parameter choice

The research question is:

**Does the advertising treatment increase the conversion rate compared with the
PSA/control group?**

Because `converted` is a binary variable, the most suitable parameter is the
conversion proportion in each group. The A/B effect is measured as:

```{r define-effect}
conversion_summary <- aggregate(
  converted ~ `test group`,
  data = marketing,
  FUN = mean
)
names(conversion_summary)[2] <- "conversion_rate"

group_counts <- aggregate(
  converted ~ `test group`,
  data = marketing,
  FUN = length
)
names(group_counts)[2] <- "n"

conversion_counts <- aggregate(
  converted ~ `test group`,
  data = marketing,
  FUN = sum
)
names(conversion_counts)[2] <- "conversions"

ab_summary <- merge(group_counts, conversion_counts, by = "test group")
ab_summary <- merge(ab_summary, conversion_summary, by = "test group")
ab_summary$conversion_percent <- 100 * ab_summary$conversion_rate

print(ab_summary)

fmt_percent <- function(x) sprintf("%.2f", 100 * x)
fmt_pvalue <- function(x) sprintf("%.4f", x)

ad_rate <- ab_summary$conversion_rate[ab_summary$`test group` == "ad"]
psa_rate <- ab_summary$conversion_rate[ab_summary$`test group` == "psa"]
observed_diff <- ad_rate - psa_rate

observed_diff
```

The observed conversion rate is `r fmt_percent(ad_rate)`% in the `ad` group
and `r fmt_percent(psa_rate)`% in the `psa` group. The observed difference is
`r fmt_percent(observed_diff)` percentage points.

# 3. Visual comparison

```{r visual-comparison}
barplot(
  ab_summary$conversion_rate,
  names.arg = ab_summary$`test group`,
  col = c("steelblue", "orange"),
  main = "Conversion Rate by Test Group",
  xlab = "Test group",
  ylab = "Conversion rate"
)
```

```{r exposure-check}
boxplot(
  `total ads` ~ `test group`,
  data = marketing,
  col = c("lightblue", "tan"),
  main = "Total Ads by Test Group",
  xlab = "Test group",
  ylab = "Total ads",
  outline = FALSE
)
```

The first plot compares the main A/B metric. The second plot is a supporting
diagnostic because the distribution of `total ads` is right-skewed and contains
large values. This makes the conversion proportion more appropriate than the
mean or variance of ad exposure for the main A/B conclusion.

# 4. Permutation test

The null hypothesis is that the treatment label has no relationship with
conversion. Under this null hypothesis, the `converted` values can be randomly
reassigned across the two groups while keeping the original group sizes fixed.

```{r permutation-test}
set.seed(42)

n_perm <- 10000
group_labels <- marketing$`test group`
converted_values <- marketing$converted

permutation_diffs <- numeric(n_perm)

for (i in seq_len(n_perm)) {
  shuffled_conversion <- sample(converted_values, replace = FALSE)

  perm_ad_rate <- mean(shuffled_conversion[group_labels == "ad"])
  perm_psa_rate <- mean(shuffled_conversion[group_labels == "psa"])

  permutation_diffs[i] <- perm_ad_rate - perm_psa_rate
}

p_value_two_sided <- mean(abs(permutation_diffs) >= abs(observed_diff))
p_value_one_sided <- mean(permutation_diffs >= observed_diff)

p_value_two_sided
p_value_one_sided
```

```{r permutation-plot}
hist(
  permutation_diffs,
  breaks = 50,
  col = "gray80",
  border = "white",
  main = "Permutation Distribution of Conversion Rate Difference",
  xlab = "Ad conversion rate - PSA conversion rate"
)
abline(v = observed_diff, col = "red", lwd = 2)
abline(v = -observed_diff, col = "red", lwd = 2, lty = 2)
```

The two-sided permutation p-value is `r fmt_pvalue(p_value_two_sided)`. The
one-sided p-value for the alternative that the ad group has a higher conversion
rate is `r fmt_pvalue(p_value_one_sided)`.

# 5. Bootstrap confidence interval

Bootstrap resampling estimates the uncertainty of the observed difference in
conversion rates. Users are resampled within each group so that the group
structure remains the same.

```{r bootstrap-test}
set.seed(42)

n_boot <- 10000
ad_converted <- marketing$converted[marketing$`test group` == "ad"]
psa_converted <- marketing$converted[marketing$`test group` == "psa"]

bootstrap_diffs <- numeric(n_boot)

for (i in seq_len(n_boot)) {
  boot_ad <- sample(ad_converted, size = length(ad_converted), replace = TRUE)
  boot_psa <- sample(psa_converted, size = length(psa_converted), replace = TRUE)

  bootstrap_diffs[i] <- mean(boot_ad) - mean(boot_psa)
}

bootstrap_ci <- quantile(bootstrap_diffs, c(0.025, 0.975))
bootstrap_ci
```

```{r bootstrap-plot}
hist(
  bootstrap_diffs,
  breaks = 50,
  col = "lightblue",
  border = "white",
  main = "Bootstrap Distribution of Conversion Rate Difference",
  xlab = "Ad conversion rate - PSA conversion rate"
)
abline(v = observed_diff, col = "red", lwd = 2)
abline(v = bootstrap_ci, col = "darkblue", lwd = 2, lty = 2)
```

The 95% bootstrap confidence interval for the difference in conversion rates is
from `r round(100 * bootstrap_ci[1], 2)` to
`r round(100 * bootstrap_ci[2], 2)` percentage points.

# 6. Comments and conclusion

* **Parameter selection:** The conversion proportion is the correct primary
  parameter because the outcome `converted` is binary. Mean, median, and
  variance are less suitable for the main A/B question because they describe
  numeric variables such as `total ads`, not the final conversion outcome.

* **Observed A/B result:** The `ad` group has a conversion rate of
  `r fmt_percent(ad_rate)`%, while the `psa` group has a conversion rate of
  `r fmt_percent(psa_rate)`%. The observed lift is
  `r fmt_percent(observed_diff)` percentage points in favor of the ad group.

* **Permutation test:** The two-sided permutation p-value is
  `r fmt_pvalue(p_value_two_sided)`. This tests whether a difference at least as
  large as the observed difference would be likely if the group labels had no
  relationship with conversion. The one-sided p-value is
  `r fmt_pvalue(p_value_one_sided)`, which directly matches the question of
  whether the ad group performs better than the PSA/control group.

* **Bootstrap interval:** The 95% bootstrap confidence interval is
  [`r round(100 * bootstrap_ci[1], 2)`,
  `r round(100 * bootstrap_ci[2], 2)`] percentage points. Because the interval is
  above zero, the mini dataset supports a positive advertising effect on
  conversion.

* **Final conclusion:** Based on the mini dataset, the advertising treatment has
  a higher conversion rate than the PSA/control condition. The permutation test
  and bootstrap confidence interval both support the conclusion that this
  observed difference is unlikely to be explained by random assignment alone.
  However, the result should still be interpreted in the context of this 10%
  sample rather than the full original dataset.
