---
title: "Marketing A/B Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(readr)
```

# Create a 10% sample dataset

The following code shows how `marketing_AB_mini.csv` was created from the
original dataset. It is shown for documentation and is not rerun as part of
this report.

```{r create-mini-dataset, eval=FALSE}
marketing_AB <- read.csv("marketing_AB.csv")

set.seed(42)

n_sample <- floor(0.10 * nrow(marketing_AB))

marketing_AB_mini <- marketing_AB[
  sample(1:nrow(marketing_AB), size = n_sample, replace = FALSE),
]

write.csv(
  marketing_AB_mini,
  "marketing_AB_mini.csv",
  row.names = FALSE
)
```

# 1. Load the data

```{r load-data}
marketing <- read_csv("marketing_AB_mini.csv")

dim(marketing)
head(marketing)
summary(marketing)

range(marketing$`most ads hour`, na.rm = TRUE)
sort(unique(marketing$`most ads hour`))
```

The dataset has `r nrow(marketing)` rows and `r ncol(marketing)` columns.

# 2. IQR outlier check

The IQR rule marks values below Q1 - 1.5 x IQR or above Q3 + 1.5 x IQR as
outliers.

```{r outlier-table}
numeric_names <- c("total ads", "most ads hour")

outlier_results <- data.frame()

for (name in numeric_names) {
  x <- marketing[[name]]
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  count <- sum(x < lower | x > upper, na.rm = TRUE)

  outlier_results <- rbind(
    outlier_results,
    data.frame(
      Variable = name,
      Q1 = q1,
      Q3 = q3,
      IQR = iqr,
      Lower_Bound = lower,
      Upper_Bound = upper,
      Outlier_Count = count,
      Contains_Outliers = ifelse(count > 0, "Yes", "No")
    )
  )
}

rownames(outlier_results) <- NULL
print(outlier_results)
```

`user id` is excluded because it is an identifier rather than a meaningful
variable for outlier analysis.

# 3. Visualizations

## Numerical distributions

```{r histograms, fig.height=4}
par(mfrow = c(1, 2))

hist(marketing$`total ads`, breaks = 50, col = "steelblue",
     main = "Distribution of Total Ads", xlab = "Total ads")

hist(marketing$`most ads hour`, breaks = seq(-0.5, 23.5, 1),
     col = "darkseagreen", main = "Distribution of Peak Ad Hour",
     xlab = "Hour")

par(mfrow = c(1, 1))
```

## Boxplot for outliers

```{r boxplot}
boxplot(marketing$`total ads`, horizontal = TRUE, col = "lightblue",
        main = "Boxplot of Total Ads", xlab = "Total ads")
```

## Categorical variables

```{r bar-charts, fig.height=7}
par(mfrow = c(2, 2), mar = c(5, 4, 3, 1))

barplot(table(marketing$`test group`), col = "tan",
        main = "Test Group", ylab = "Number of users")

barplot(table(marketing$converted), col = "skyblue",
        main = "Conversion", ylab = "Number of users")

barplot(table(marketing$`most ads day`), col = "plum",
        main = "Peak Ad Day", ylab = "Number of users", las = 2)

par(mfrow = c(1, 1))
```

## Advertising exposure and conversion

```{r exposure-conversion}
boxplot(`total ads` ~ converted, data = marketing,
        names = c("Not converted", "Converted"),
        col = c("lightgray", "lightgreen"),
        main = "Ad Exposure by Conversion",
        xlab = "Conversion", ylab = "Total ads",
        outline = FALSE)
```

```{r exposure-groups}
marketing$exposure_group <- cut(
  marketing$`total ads`,
  breaks = c(0, 5, 10, 20, 40, 80, Inf),
  labels = c("1-5", "6-10", "11-20", "21-40", "41-80", "81+")
)

conversion_by_exposure <- aggregate(
  converted ~ exposure_group,
  data = marketing,
  FUN = mean
)

plot(seq_len(nrow(conversion_by_exposure)),
     conversion_by_exposure$converted,
     type = "b", pch = 19, col = "darkblue",
     xaxt = "n",
     main = "Conversion Rate by Ad Exposure",
     xlab = "Total ad exposure group", ylab = "Conversion rate")
axis(1, at = seq_len(nrow(conversion_by_exposure)),
     labels = conversion_by_exposure$exposure_group)
```

## Test group and conversion

```{r group-conversion}
group_summary <- aggregate(
  converted ~ `test group`,
  data = marketing,
  FUN = mean
)
names(group_summary)[2] <- "conversion_rate"

print(group_summary)

barplot(group_summary$conversion_rate,
        names.arg = group_summary$`test group`,
        col = c("steelblue", "orange"),
        main = "Conversion Rate by Test Group",
        xlab = "Test group", ylab = "Conversion rate")
```

## Conversion patterns by day and hour

```{r timing-summary}
day_summary <- aggregate(
  converted ~ `most ads day`, data = marketing, FUN = mean
)
names(day_summary)[2] <- "conversion_rate"
day_summary <- day_summary[order(day_summary$conversion_rate,
                                 decreasing = TRUE), ]
print(day_summary)

hour_summary <- aggregate(
  converted ~ `most ads hour`, data = marketing, FUN = mean
)
names(hour_summary)[2] <- "conversion_rate"
print(hour_summary)

barplot(day_summary$conversion_rate,
        names.arg = day_summary$`most ads day`,
        main = "Conversion Rate by Most Ads Day",
        xlab = "Most Ads Day", ylab = "Conversion Rate", las = 2)

plot(hour_summary$`most ads hour`,
     hour_summary$conversion_rate,
     type = "b", pch = 19,
     main = "Conversion Rate by Most Ads Hour",
     xlab = "Most Ads Hour", ylab = "Conversion Rate")
```

# 4. Comments and observations

* **Distribution of numerical variables:** Total ad exposure is strongly
  right-skewed. The median user saw `r median(marketing[["total ads"]])` ads,
  while the mean was `r round(mean(marketing[["total ads"]]), 1)` ads. The mean
  is higher because a small number of users received very large numbers of ads;
  the maximum was `r max(marketing[["total ads"]])`. This suggests that the
  typical user had modest exposure, but a smaller group remained exposed to the campaign long enough to receive ads repeatedly. The range check shows that the observed `most ads hour` values run from 0 to 23. Peak exposure was most
  common at hour
  `r as.numeric(names(which.max(table(marketing[["most ads hour"]]))))`, showing
  that exposure volume was concentrated around midday.

* **Outliers:** For `total ads`, Q1 is
  `r outlier_results$Q1[outlier_results$Variable == "total ads"]`, Q3 is
  `r outlier_results$Q3[outlier_results$Variable == "total ads"]`, and the upper
  IQR boundary is
  `r outlier_results$Upper_Bound[outlier_results$Variable == "total ads"]`.
  Users with more than 61.5 ads are therefore flagged, producing
  `r outlier_results$Outlier_Count[outlier_results$Variable == "total ads"]`
  outliers. They may represent highly engaged users, users with longer campaign exposure,
  or repeated targeting; they should not automatically be treated as data errors. The method also flags
  `r outlier_results$Outlier_Count[outlier_results$Variable == "most ads hour"]`
  records for `most ads hour` because its IQR lower bound is 0.5. In this data,
  the flagged values are hour 0. The range and unique-value checks confirm that hour 0 is an observed timing
  value, so it should not automatically be treated as invalid or as a data error. It may simply be less common than
  hours near the center of the distribution. Because `most ads hour` is a
  time-related, descriptive variable rather than a continuous quantity, its IQR
  result requires careful interpretation. The main meaningful outlier finding
  is the high-value tail of `total ads`, which directly measures how many ads
  users saw.

* **Advertising exposure and conversion:** Converted users saw an average of
  `r round(mean(marketing[["total ads"]][marketing$converted]), 1)` ads and a
  median of `r median(marketing[["total ads"]][marketing$converted])`, compared
  with an average of
  `r round(mean(marketing[["total ads"]][!marketing$converted]), 1)` and a median
  of `r median(marketing[["total ads"]][!marketing$converted])` for users who did
  not convert. Conversion increased from about
  `r round(100 * conversion_by_exposure$converted[1], 2)`% among users seeing
  1-5 ads to `r round(100 * tail(conversion_by_exposure$converted, 1), 2)`% among
  those seeing 81 or more. Higher exposure is strongly associated with higher
  conversion. It does not prove that more ads alone cause conversion because
  users who stay active longer have more chances both to see ads and to convert.

* **Conversion behavior and test groups:** Only
  `r sum(marketing$converted)` of `r nrow(marketing)` users converted, an overall
  rate of `r round(100 * mean(marketing$converted), 2)`%. The `ad` group conversion rate was
  `r round(100 * group_summary[group_summary[["test group"]] == "ad", "conversion_rate"], 2)`%,
  compared with
  `r round(100 * group_summary[group_summary[["test group"]] == "psa", "conversion_rate"], 2)`%
  in the PSA/control group. The ad-group rate is about
  `r round(group_summary[group_summary[["test group"]] == "ad", "conversion_rate"] / group_summary[group_summary[["test group"]] == "psa", "conversion_rate"], 2)`
  times the control rate. This observed difference suggests that the ad treatment performed better than
  the PSA/control treatment in this sample. Group sizes are very unequal, so rates are more useful than raw
  conversion counts for this comparison.

* **Timing patterns:** Friday had the most users whose peak exposure occurred
  on that day, but it did not have the highest conversion rate. Monday had the
  highest day-level rate at
  `r round(100 * day_summary$conversion_rate[day_summary[["most ads day"]] == "Monday"], 2)`%,
  while Friday was lowest at
  `r round(100 * day_summary$conversion_rate[day_summary[["most ads day"]] == "Friday"], 2)`%.
  Exposure volume was concentrated around hours 11-15. Hour 16 had a relatively
  high conversion rate of
  `r round(100 * hour_summary$conversion_rate[hour_summary[["most ads hour"]] == 16], 2)`%
  and also had many users. Hour 6 had the highest raw hourly rate, but only
  `r sum(marketing[["most ads hour"]] == 6)` users were in that segment, so its
  rate is less stable. These descriptive differences may reflect audience behavior or campaign
  scheduling, rather than an effect of timing alone.
