---
title: "R Lecture: Writing Pipeline-Structured Functions"
author: "TS.To Duc Khanh"
date: "17/05/2026"
output: html_document
---

In R data processing, a "pipeline-friendly" function typically has these characteristics:

- The first argument is the data (`.data`)
- Uses the `%>%` pipe or the native `|>` pipe
- Returns a `tibble`/`data.frame`
- Supports non-standard evaluation (NSE) with `{{ }}` or `.data`
- Compatible with `dplyr`, `purrr`, `tidyr`, `ggplot2`

The example below illustrates how to write a standard `tidyverse` function that:

- groups data,
- computes descriptive statistics,
- filters by a threshold,
- and returns a tibble.

```{r, message=FALSE, warning=FALSE}
library(dplyr)

summarise_by_group <- function(.data, group_var, target_var,
                               threshold = NULL, ...) {
  result <- .data |>
    group_by({{ group_var }}) |>
    summarise(
      mean = mean({{ target_var }}, ...),
      sd   = sd({{ target_var }}, ...),
      n    = n(),
      .groups = "drop"
    )

  if (!is.null(threshold)) {
    result <- result |>
      filter(mean >= threshold)
  }

  result
}

iris |> summarise_by_group(Species, Sepal.Length)

iris |> summarise_by_group(Species, Sepal.Length, threshold = 6)
```

In the pipeline function above, the arguments are interpreted as follows:

- `.data` is always the first argument, which allows the function to work well with pipes
- `{{ }}` enables variable name look-up following the tidy evaluation standard

---

**Exercise 1:** Write a pipeline function that standardizes a variable:

$$y = \frac{x - \bar{x}}{s},$$

where $\bar{x}$ and $s$ are the sample mean and sample standard deviation of the variable to be standardized.

```{r}
standardize <- function(.data, target_var, ...) {
  col_name <- paste0(deparse(substitute(target_var)), "_scaled")
  x        <- pull(.data, {{ target_var }})
  mean_x   <- mean(x, na.rm = TRUE)
  sd_x     <- sd(x, na.rm = TRUE)
  result <- .data |>
    mutate(.scaled = ({{ target_var }} - mean_x) / sd_x) |>
    as_tibble()
  names(result)[names(result) == ".scaled"] <- col_name
  result
}

iris |>
  standardize(Sepal.Length) |>
  select(Species, Sepal.Length, Sepal.Length_scaled) |>
  head()
```

---

**Exercise 2:** Write a pipeline function that identifies outliers using the Hampel filter.

The Hampel filter detects outliers based on the **Median Absolute Deviation (MAD)**:

$$\text{MAD} = \text{median}(|x_i - \text{median}(x)|)$$

An observation $x_i$ is flagged as an outlier if:

$$|x_i - \text{median}(x)| > k \times 1.4826 \times \text{MAD}$$

where $k = 3$ by default (the constant $1.4826$ makes MAD consistent with the standard deviation under a normal distribution).

```{r}
detect_outliers_hampel <- function(.data, target_var, k = 3, ...) {
  x     <- pull(.data, {{ target_var }})
  med_x <- median(x, na.rm = TRUE)
  mad_x <- mad(x, na.rm = TRUE)
  .data |>
    mutate(is_outlier = abs({{ target_var }} - med_x) > k * mad_x) |>
    as_tibble()
}

iris |>
  detect_outliers_hampel(Sepal.Length) |>
  filter(is_outlier == TRUE)

iris |>
  detect_outliers_hampel(Sepal.Length) |>
  count(Species, is_outlier)
```

**Note:** Some principles to keep in mind:

- Do not use `attach()`
- Use `.data[[ ]]` when a variable name is passed as a string
- Prefer `tibble` over `data.frame`
