---
title: "Mean Shift Clustering - Experimental Analysis"
author: "Quang Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    css: ../../templates/styles.css
    toc: true
    toc_float: true
    number_sections: true
    theme: default
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load required libraries
library(ggplot2)
library(cluster)
library(mclust)
library(gridExtra)
library(dplyr)

# Source algorithm implementations
source("mean_shift_clustering_algorithm.R")
source("mean_shift_clustering_helpers.R")
```

# Introduction

This document presents experimental results for the Mean Shift Clustering algorithm. We focus on three key experiments:

1. **Varying Data Distributions**: Testing on different numbers of points and cluster densities
2. **Kernel Function Comparison**: Comparing Gaussian, Flat, and Epanechnikov kernels
3. **Bandwidth Sensitivity**: Analyzing the critical impact of bandwidth parameter selection

# Experiment Setup

## Data Generation Function

```{r}
#' Generate synthetic 2D blobs for clustering
#'
#' @param n_samples Total number of samples
#' @param centers Matrix of cluster centers
#' @param cluster_std Standard deviation of each cluster
#' @return List with data and true labels
generate_blobs <- function(n_samples, centers, cluster_std = 0.5) {
  n_centers <- nrow(centers)
  samples_per_cluster <- rep(n_samples %/% n_centers, n_centers)
  samples_per_cluster[1] <- samples_per_cluster[1] + (n_samples %% n_centers)

  data_list <- list()
  labels <- c()

  for (i in 1:n_centers) {
    n <- samples_per_cluster[i]
    cluster_data <- matrix(rnorm(n * 2, mean = 0, sd = cluster_std), ncol = 2)
    cluster_data[, 1] <- cluster_data[, 1] + centers[i, 1]
    cluster_data[, 2] <- cluster_data[, 2] + centers[i, 2]

    data_list[[i]] <- cluster_data
    labels <- c(labels, rep(i, n))
  }

  data <- do.call(rbind, data_list)
  list(data = data, labels = labels)
}

#' Evaluate clustering performance
#'
#' @param data Data matrix
#' @param result Mean shift result
#' @param true_labels Ground truth labels
#' @return Data frame with metrics
evaluate_clustering <- function(data, result, true_labels) {
  silhouette <- compute_silhouette(data, result$cluster_labels)
  ari <- compute_ari(true_labels, result$cluster_labels)

  data.frame(
    n_clusters = result$n_clusters,
    silhouette = round(silhouette, 3),
    ari = round(ari, 3),
    time_secs = round(result$computation_time, 3),
    mean_iterations = round(result$mean_iterations, 2)
  )
}
```

# Experiment 1: Varying Data Distributions

**Objective**: Test Mean Shift on different blob configurations to understand how data characteristics affect performance.

## Setup

We test three configurations:
- **Sparse**: 100 points, 3 well-separated clusters
- **Medium**: 500 points, 3 moderately separated clusters
- **Dense**: 1000 points, 3 overlapping clusters

```{r exp1_setup}
# Set seed for reproducibility
set.seed(42)

# Configuration 1: Sparse (100 points)
sparse_data <- generate_blobs(
  n_samples = 100,
  centers = matrix(c(0, 0, 5, 5, 10, 0), ncol = 2, byrow = TRUE),
  cluster_std = 0.5
)

# Configuration 2: Medium (500 points)
medium_data <- generate_blobs(
  n_samples = 500,
  centers = matrix(c(0, 0, 5, 5, 10, 0), ncol = 2, byrow = TRUE),
  cluster_std = 0.5
)

# Configuration 3: Dense overlapping (1000 points)
dense_data <- generate_blobs(
  n_samples = 1000,
  centers = matrix(c(0, 0, 3, 3, 6, 0), ncol = 2, byrow = TRUE),
  cluster_std = 0.8
)
```

## Visualization: Data Configurations

```{r exp1_viz_data, fig.height=4}
# Plot all three configurations
p1 <- ggplot(data.frame(x = sparse_data$data[, 1],
                        y = sparse_data$data[, 2],
                        cluster = factor(sparse_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 2) +
  labs(title = "Sparse (100 pts)", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(data.frame(x = medium_data$data[, 1],
                        y = medium_data$data[, 2],
                        cluster = factor(medium_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  labs(title = "Medium (500 pts)", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(data.frame(x = dense_data$data[, 1],
                        y = dense_data$data[, 2],
                        cluster = factor(dense_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1) +
  labs(title = "Dense overlapping (1000 pts)", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, ncol = 3)
```

## Run Mean Shift Clustering

```{r exp1_run}
# Run Mean Shift on each configuration with optimal bandwidth
result_sparse <- mean_shift_clustering(
  sparse_data$data,
  bandwidth = 0.8,
  kernel = "gaussian"
)

result_medium <- mean_shift_clustering(
  medium_data$data,
  bandwidth = 0.8,
  kernel = "gaussian"
)

result_dense <- mean_shift_clustering(
  dense_data$data,
  bandwidth = 0.8,
  kernel = "gaussian"
)
```

## Results Table

```{r exp1_results}
results_exp1 <- rbind(
  cbind(Configuration = "Sparse (100 pts)",
        evaluate_clustering(sparse_data$data, result_sparse, sparse_data$labels)),
  cbind(Configuration = "Medium (500 pts)",
        evaluate_clustering(medium_data$data, result_medium, medium_data$labels)),
  cbind(Configuration = "Dense (1000 pts)",
        evaluate_clustering(dense_data$data, result_dense, dense_data$labels))
)

knitr::kable(results_exp1, caption = "Experiment 1: Varying Data Distributions")
```

## Visualization: Clustering Results

```{r exp1_viz_results, fig.height=4}
p1_result <- ggplot(data.frame(x = sparse_data$data[, 1],
                               y = sparse_data$data[, 2],
                               cluster = factor(result_sparse$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 2) +
  geom_point(data = data.frame(result_sparse$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Sparse: %d clusters", result_sparse$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p2_result <- ggplot(data.frame(x = medium_data$data[, 1],
                               y = medium_data$data[, 2],
                               cluster = factor(result_medium$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_medium$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Medium: %d clusters", result_medium$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p3_result <- ggplot(data.frame(x = dense_data$data[, 1],
                               y = dense_data$data[, 2],
                               cluster = factor(result_dense$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1) +
  geom_point(data = data.frame(result_dense$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Dense: %d clusters", result_dense$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1_result, p2_result, p3_result, ncol = 3)
```

## Key Observations

- **Performance degrades with overlapping clusters**: Dense configuration shows lower Silhouette and ARI scores
- **Computation time scales with data size**: Time increases from sparse to dense as expected with O(n²) complexity
- **Extra clusters on dense data**: The algorithm may find more clusters than true clusters when data overlaps

# Experiment 2: Kernel Function Comparison

**Objective**: Compare the performance of different kernel functions on the same dataset.

## Setup

Using the medium dataset (500 points, 3 clusters), we compare:
- Gaussian kernel
- Flat (Uniform) kernel
- Epanechnikov kernel

```{r exp2_run}
# Use medium dataset with optimal bandwidth
bandwidth_optimal <- 0.8

result_gaussian <- mean_shift_clustering(
  medium_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "gaussian"
)

result_flat <- mean_shift_clustering(
  medium_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "flat"
)

result_epanechnikov <- mean_shift_clustering(
  medium_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "epanechnikov"
)
```

## Results Table

```{r exp2_results}
results_exp2 <- rbind(
  cbind(Kernel = "Gaussian",
        evaluate_clustering(medium_data$data, result_gaussian, medium_data$labels)),
  cbind(Kernel = "Flat (Uniform)",
        evaluate_clustering(medium_data$data, result_flat, medium_data$labels)),
  cbind(Kernel = "Epanechnikov",
        evaluate_clustering(medium_data$data, result_epanechnikov, medium_data$labels))
)

knitr::kable(results_exp2, caption = "Experiment 2: Kernel Function Comparison")
```

## Visualization: Kernel Comparison

```{r exp2_viz, fig.height=4}
p_gaussian <- ggplot(data.frame(x = medium_data$data[, 1],
                                y = medium_data$data[, 2],
                                cluster = factor(result_gaussian$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_gaussian$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Gaussian Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_flat <- ggplot(data.frame(x = medium_data$data[, 1],
                            y = medium_data$data[, 2],
                            cluster = factor(result_flat$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_flat$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Flat (Uniform) Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_epan <- ggplot(data.frame(x = medium_data$data[, 1],
                            y = medium_data$data[, 2],
                            cluster = factor(result_epanechnikov$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_epanechnikov$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Epanechnikov Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_gaussian, p_flat, p_epan, ncol = 3)
```

## Key Findings

- **Flat kernel is fastest**: Offers best speed-accuracy trade-off
- **Results are generally similar**: All three kernels produce comparable clustering quality
- **Kernel choice matters less than bandwidth**: The bandwidth parameter has more impact on results

# Experiment 3: Bandwidth Sensitivity

**Objective**: Demonstrate the critical importance of bandwidth parameter selection.

## Setup

Using the medium dataset, we test three bandwidth values:
- **Small**: h = 0.3 (too small, over-segmentation expected)
- **Optimal**: h = 0.8 (appropriate for this dataset)
- **Large**: h = 2.0 (too large, under-segmentation expected)

```{r exp3_run}
result_small <- mean_shift_clustering(
  medium_data$data,
  bandwidth = 0.3,
  kernel = "gaussian"
)

result_optimal <- mean_shift_clustering(
  medium_data$data,
  bandwidth = 0.8,
  kernel = "gaussian"
)

result_large <- mean_shift_clustering(
  medium_data$data,
  bandwidth = 2.0,
  kernel = "gaussian"
)
```

## Results Table

```{r exp3_results}
results_exp3 <- rbind(
  cbind(Bandwidth = "Small (h = 0.3)",
        evaluate_clustering(medium_data$data, result_small, medium_data$labels)),
  cbind(Bandwidth = "Optimal (h = 0.8)",
        evaluate_clustering(medium_data$data, result_optimal, medium_data$labels)),
  cbind(Bandwidth = "Large (h = 2.0)",
        evaluate_clustering(medium_data$data, result_large, medium_data$labels))
)

knitr::kable(results_exp3, caption = "Experiment 3: Bandwidth Sensitivity")
```

## Visualization: Bandwidth Impact

```{r exp3_viz, fig.height=4}
p_small <- ggplot(data.frame(x = medium_data$data[, 1],
                             y = medium_data$data[, 2],
                             cluster = factor(result_small$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_small$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Small h=0.3: %d clusters", result_small$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_optimal <- ggplot(data.frame(x = medium_data$data[, 1],
                               y = medium_data$data[, 2],
                               cluster = factor(result_optimal$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_optimal$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Optimal h=0.8: %d clusters", result_optimal$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_large <- ggplot(data.frame(x = medium_data$data[, 1],
                             y = medium_data$data[, 2],
                             cluster = factor(result_large$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_large$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Large h=2.0: %d clusters", result_large$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_small, p_optimal, p_large, ncol = 3)
```

## Bandwidth vs. Cluster Count Analysis

```{r exp3_bandwidth_analysis, fig.height=5}
# Test range of bandwidth values
bandwidth_values <- seq(0.2, 3.0, by = 0.1)
n_clusters_found <- numeric(length(bandwidth_values))

for (i in seq_along(bandwidth_values)) {
  result_temp <- mean_shift_clustering(
    medium_data$data,
    bandwidth = bandwidth_values[i],
    kernel = "gaussian"
  )
  n_clusters_found[i] <- result_temp$n_clusters
}

# Plot
ggplot(data.frame(bandwidth = bandwidth_values, clusters = n_clusters_found)) +
  geom_line(aes(x = bandwidth, y = clusters), size = 1, color = "blue") +
  geom_point(aes(x = bandwidth, y = clusters), size = 2) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 2.5, y = 3.5, label = "True clusters = 3", color = "red") +
  geom_vline(xintercept = 0.8, linetype = "dashed", color = "darkgreen", size = 1) +
  annotate("text", x = 1.2, y = max(n_clusters_found) - 1,
           label = "Optimal h = 0.8", color = "darkgreen") +
  labs(title = "Bandwidth vs. Number of Clusters Found",
       x = "Bandwidth (h)",
       y = "Number of Clusters") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

## Critical Insights

- **Too small bandwidth → over-segmentation**: Many tiny clusters instead of true 3 clusters
- **Too large bandwidth → under-segmentation**: All points merge into 1 cluster
- **Bandwidth is the most critical parameter**: More important than kernel choice
- **Automatic selection methods**: Scott's and Silverman's rules provide good starting points

# Bandwidth Selection Strategies

```{r bandwidth_strategies}
# Compare automatic bandwidth selection methods
bandwidth_scott <- scotts_rule(medium_data$data)
bandwidth_silverman <- silvermans_rule(medium_data$data)

cat(sprintf("Scott's Rule: h = %.4f\n", bandwidth_scott))
cat(sprintf("Silverman's Rule: h = %.4f\n", bandwidth_silverman))
cat(sprintf("Manually chosen optimal: h = 0.8\n"))

# Test both automatic methods
result_scott <- mean_shift_clustering(
  medium_data$data,
  bandwidth = bandwidth_scott,
  kernel = "gaussian"
)

result_silverman <- mean_shift_clustering(
  medium_data$data,
  bandwidth = bandwidth_silverman,
  kernel = "gaussian"
)

bandwidth_strategies <- rbind(
  cbind(Method = "Scott's Rule", h = round(bandwidth_scott, 3),
        evaluate_clustering(medium_data$data, result_scott, medium_data$labels)),
  cbind(Method = "Silverman's Rule", h = round(bandwidth_silverman, 3),
        evaluate_clustering(medium_data$data, result_silverman, medium_data$labels)),
  cbind(Method = "Manual Optimal", h = 0.8,
        evaluate_clustering(medium_data$data, result_optimal, medium_data$labels))
)

knitr::kable(bandwidth_strategies, caption = "Bandwidth Selection Strategies Comparison")
```

# Summary of Experimental Findings

## Key Takeaways

1. **Data Characteristics**:
   - Works well on well-separated clusters
   - Struggles with heavy overlap
   - Computation time scales with data size (O(n²) complexity)

2. **Kernel Choice**:
   - Flat/Uniform kernel: fastest, good results
   - Gaussian: slightly better quality, slower
   - Choice matters less than bandwidth selection

3. **Bandwidth Selection** (MOST CRITICAL):
   - Most important parameter affecting results
   - Use adaptive methods (Scott's/Silverman's rule) as starting point
   - Fine-tune based on domain knowledge or cross-validation
   - Validate with silhouette score

## Practical Recommendations

- Start with automatic bandwidth selection (Scott's or Silverman's rule)
- Visualize results and adjust if needed
- For well-separated clusters, use slightly larger bandwidth
- For overlapping data, use smaller bandwidth but expect more clusters
- Consider computational cost for large datasets (n > 10,000)

# Session Info

```{r session_info}
sessionInfo()
```
