---
title: "Mean_shift_clustering - Documentation"
author: "Quang Nguyen"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: ../../templates/styles.css
    toc: true
    toc_float: true
    number_sections: true
    theme: default
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Algorithm Overview

Brief introduction to the algorithm, its purpose, and when to use it.

## Background

Describe the theoretical background and motivation for this algorithm.

## Key Concepts

Explain the main concepts and principles behind the algorithm.

# Mathematical Foundation

## Theory

Present the mathematical formulation and theoretical foundations.

## Algorithm Steps

Describe the algorithm step-by-step:

1. **Step 1**: Description
2. **Step 2**: Description
3. **Step 3**: Description

# Implementation

## Main Function

```{r}
# Source the algorithm implementation
source("mean_shift_clustering.R")

# Example function signature
# mean_shift_clustering(data, parameters)
```

## Helper Functions

```{r}
# Source helper functions
source("algorithm_helpers.R")
```

# Examples

## Basic Example

```{r}
# Load required libraries
library(ggplot2)

# Example code here
# data <- ...
# result <- mean_shift_clustering(data, ...)
# plot(result)
```

## Advanced Example

```{r}
# More complex example
```

# Performance Analysis

## Complexity

Discuss time and space complexity.

## Benchmarks

```{r}
# Performance benchmarks
```

# References

- Reference 1
- Reference 2
- Reference 3

# Appendix

## Additional Notes

Any additional information, edge cases, or implementation details.

