---
title: "Mean Shift Clustering - Experimental Analysis"
author: "Quang Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    css: ../../templates/styles.css
    toc: true
    toc_float: true
    number_sections: true
    theme: default
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load required libraries
library(ggplot2)
library(cluster)
library(mclust)
library(gridExtra)
library(dplyr)

# Source algorithm implementations
source("mean_shift_clustering_algorithm.R")
source("mean_shift_clustering_helpers.R")
```

# Introduction

This document presents experimental results for the Mean Shift Clustering algorithm. We focus on four key experiments:

1. **Varying Data Distributions**: Testing on different numbers of points and cluster densities
2. **Kernel Function Comparison**: Comparing Gaussian, Flat, and Epanechnikov kernels
3. **Bandwidth Sensitivity**: Analyzing the critical impact of bandwidth parameter selection
4. **Scalability Analysis**: Runtime performance as dataset size increases

# Experiment Setup

## Data Generation Function

```{r}
#' Generate synthetic 2D blobs for clustering
#'
#' @param n_samples Total number of samples
#' @param centers Matrix of cluster centers
#' @param cluster_std Standard deviation of each cluster
#' @return List with data and true labels
generate_blobs <- function(n_samples, centers, cluster_std = 0.7) {
  n_centers <- nrow(centers)
  samples_per_cluster <- rep(n_samples %/% n_centers, n_centers)
  samples_per_cluster[1] <- samples_per_cluster[1] + (n_samples %% n_centers)

  data_list <- list()
  labels <- c()

  for (i in 1:n_centers) {
    n <- samples_per_cluster[i]
    cluster_data <- matrix(rnorm(n * 2, mean = 0, sd = cluster_std), ncol = 2)
    cluster_data[, 1] <- cluster_data[, 1] + centers[i, 1]
    cluster_data[, 2] <- cluster_data[, 2] + centers[i, 2]

    data_list[[i]] <- cluster_data
    labels <- c(labels, rep(i, n))
  }

  data <- do.call(rbind, data_list)
  list(data = data, labels = labels)
}

#' Evaluate clustering performance
#'
#' @param data Data matrix
#' @param result Mean shift result
#' @param true_labels Ground truth labels
#' @return Data frame with metrics
evaluate_clustering <- function(data, result, true_labels) {
  silhouette <- compute_silhouette(data, result$cluster_labels)
  ari <- compute_ari(true_labels, result$cluster_labels)

  data.frame(
    n_clusters = result$n_clusters,
    silhouette = round(silhouette, 3),
    ari = round(ari, 3),
    time_secs = round(result$computation_time, 3),
    mean_iterations = round(result$mean_iterations, 2)
  )
}
```

# Experiment 1: Varying Data Distributions

**Objective**: Test Mean Shift on datasets with varying levels of cluster overlap to understand how the density-based algorithm handles different separation scenarios.

## Setup

We test three configurations with the same sample size (500 points, 3 clusters) but different overlap levels:
- **Well-separated**: Clear gaps between clusters
- **Moderate overlap**: Some overlap at cluster boundaries
- **Heavy overlap**: Significant overlap, challenging for any clustering algorithm

```{r exp1_setup}
# Set seed for reproducibility
set.seed(42)

# Configuration 1: Well-separated (500 points)
well_separated_data <- generate_blobs(
  n_samples = 500,
  centers = matrix(c(0, 0, 5, 5, 10, 0), ncol = 2, byrow = TRUE),
  cluster_std = 0.6
)

# Configuration 2: Moderate overlap (500 points)
moderate_overlap_data <- generate_blobs(
  n_samples = 500,
  centers = matrix(c(0, 0, 3, 3, 6, 0), ncol = 2, byrow = TRUE),
  cluster_std = 0.8
)

# Configuration 3: Heavy overlap (500 points)
heavy_overlap_data <- generate_blobs(
  n_samples = 500,
  centers = matrix(c(0, 0, 2, 2, 4, 0), ncol = 2, byrow = TRUE),
  cluster_std = 1.0
)
```

## Visualization: Data Configurations

```{r exp1_viz_data, fig.height=4}
# Plot all three configurations
p1 <- ggplot(data.frame(x = well_separated_data$data[, 1],
                        y = well_separated_data$data[, 2],
                        cluster = factor(well_separated_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  labs(title = "Well-separated", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(data.frame(x = moderate_overlap_data$data[, 1],
                        y = moderate_overlap_data$data[, 2],
                        cluster = factor(moderate_overlap_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  labs(title = "Moderate overlap", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(data.frame(x = heavy_overlap_data$data[, 1],
                        y = heavy_overlap_data$data[, 2],
                        cluster = factor(heavy_overlap_data$labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  labs(title = "Heavy overlap", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, ncol = 3)
```

## Run Mean Shift Clustering

```{r exp1_run}
# Run Mean Shift on each configuration with flat kernel
result_well_separated <- mean_shift_clustering(
  well_separated_data$data,
  bandwidth = 1.2,
  kernel = "flat"
)

result_moderate <- mean_shift_clustering(
  moderate_overlap_data$data,
  bandwidth = 1.2,
  kernel = "flat"
)

result_heavy <- mean_shift_clustering(
  heavy_overlap_data$data,
  bandwidth = 1.2,
  kernel = "flat"
)
```

## Results Table

```{r exp1_results}
results_exp1 <- rbind(
  cbind(Configuration = "Well-separated",
        evaluate_clustering(well_separated_data$data, result_well_separated, well_separated_data$labels)),
  cbind(Configuration = "Moderate overlap",
        evaluate_clustering(moderate_overlap_data$data, result_moderate, moderate_overlap_data$labels)),
  cbind(Configuration = "Heavy overlap",
        evaluate_clustering(heavy_overlap_data$data, result_heavy, heavy_overlap_data$labels))
)

knitr::kable(results_exp1, caption = "Experiment 1: Effect of Cluster Overlap on Mean Shift Performance")
```

## Visualization: Clustering Results

```{r exp1_viz_results, fig.height=4}
p1_result <- ggplot(data.frame(x = well_separated_data$data[, 1],
                               y = well_separated_data$data[, 2],
                               cluster = factor(result_well_separated$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_well_separated$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Well-separated: %d clusters", result_well_separated$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p2_result <- ggplot(data.frame(x = moderate_overlap_data$data[, 1],
                               y = moderate_overlap_data$data[, 2],
                               cluster = factor(result_moderate$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_moderate$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Moderate: %d clusters", result_moderate$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p3_result <- ggplot(data.frame(x = heavy_overlap_data$data[, 1],
                               y = heavy_overlap_data$data[, 2],
                               cluster = factor(result_heavy$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_heavy$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Heavy: %d clusters", result_heavy$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1_result, p2_result, p3_result, ncol = 3)
```

## Key Observations

- **Well-separated clusters**: Algorithm easily identifies correct number of clusters with high Silhouette and ARI scores
- **Moderate overlap**: Performance remains reasonable, but some boundary points may be misclassified
- **Heavy overlap**: Significant performance degradation as density peaks merge or fragment
- **Flat kernel advantage**: Fast computation while maintaining good clustering quality

# Experiment 2: Kernel Function Comparison

**Objective**: Compare the performance of different kernel functions on the same dataset.

## Setup

Using the dense dataset (1000 points, 3 clusters), we compare:
- Gaussian kernel
- Flat (Uniform) kernel
- Epanechnikov kernel

```{r exp2_run}
# Use medium dataset with optimal bandwidth
bandwidth_optimal <- 1.2

result_gaussian <- mean_shift_clustering(
  dense_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "gaussian"
)

result_flat <- mean_shift_clustering(
  dense_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "flat"
)

result_epanechnikov <- mean_shift_clustering(
  dense_data$data,
  bandwidth = bandwidth_optimal,
  kernel = "epanechnikov"
)
```

## Results Table

```{r exp2_results}
results_exp2 <- rbind(
  cbind(Kernel = "Gaussian",
        evaluate_clustering(dense_data$data, result_gaussian, dense_data$labels)),
  cbind(Kernel = "Flat (Uniform)",
        evaluate_clustering(dense_data$data, result_flat, dense_data$labels)),
  cbind(Kernel = "Epanechnikov",
        evaluate_clustering(dense_data$data, result_epanechnikov, dense_data$labels))
)

knitr::kable(results_exp2, caption = "Experiment 2: Kernel Function Comparison")
```

## Visualization: Kernel Comparison

```{r exp2_viz, fig.height=4}
p_gaussian <- ggplot(data.frame(x = dense_data$data[, 1],
                                y = dense_data$data[, 2],
                                cluster = factor(result_gaussian$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_gaussian$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Gaussian Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_flat <- ggplot(data.frame(x = dense_data$data[, 1],
                            y = dense_data$data[, 2],
                            cluster = factor(result_flat$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_flat$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Flat (Uniform) Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_epan <- ggplot(data.frame(x = dense_data$data[, 1],
                            y = dense_data$data[, 2],
                            cluster = factor(result_epanechnikov$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_epanechnikov$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = "Epanechnikov Kernel", x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_gaussian, p_flat, p_epan, ncol = 3)
```

## Key Findings

- **Flat kernel is fastest**: Offers best speed-accuracy trade-off
- **Results are generally similar**: All three kernels produce comparable clustering quality
- **Kernel choice matters less than bandwidth**: The bandwidth parameter has more impact on results

# Experiment 3: Bandwidth Sensitivity

**Objective**: Demonstrate the critical importance of bandwidth parameter selection.

## Setup

Using the medium dataset, we test three bandwidth values:
- **Small**: h = 0.3 (too small, over-segmentation expected)
- **Optimal**: h = 0.8 (appropriate for this dataset)
- **Large**: h = 2.0 (too large, under-segmentation expected)

```{r exp3_run}
result_small <- mean_shift_clustering(
  dense_data$data,
  bandwidth = 0.3,
  kernel = "gaussian"
)

result_optimal <- mean_shift_clustering(
  dense_data$data,
  bandwidth = 1.0,
  kernel = "gaussian"
)

result_large <- mean_shift_clustering(
  dense_data$data,
  bandwidth = 2.0,
  kernel = "gaussian"
)
```

## Results Table

```{r exp3_results}
results_exp3 <- rbind(
  cbind(Bandwidth = "Small (h = 0.3)",
        evaluate_clustering(dense_data$data, result_small, dense_data$labels)),
  cbind(Bandwidth = "Optimal (h = 1.0)",
        evaluate_clustering(dense_data$data, result_optimal, dense_data$labels)),
  cbind(Bandwidth = "Large (h = 2.0)",
        evaluate_clustering(dense_data$data, result_large, dense_data$labels))
)

knitr::kable(results_exp3, caption = "Experiment 3: Bandwidth Sensitivity")
```

## Visualization: Bandwidth Impact

```{r exp3_viz, fig.height=4}
p_small <- ggplot(data.frame(x = dense_data$data[, 1],
                             y = dense_data$data[, 2],
                             cluster = factor(result_small$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_small$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Small h=0.3: %d clusters", result_small$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_optimal <- ggplot(data.frame(x = dense_data$data[, 1],
                               y = dense_data$data[, 2],
                               cluster = factor(result_optimal$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_optimal$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Optimal h=1.0: %d clusters", result_optimal$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

p_large <- ggplot(data.frame(x = dense_data$data[, 1],
                             y = dense_data$data[, 2],
                             cluster = factor(result_large$cluster_labels))) +
  geom_point(aes(x = x, y = y, color = cluster), size = 1.5) +
  geom_point(data = data.frame(result_large$modes),
             aes(x = X1, y = X2), size = 5, shape = 4, stroke = 2, color = "black") +
  labs(title = sprintf("Large h=2.0: %d clusters", result_large$n_clusters),
       x = "X", y = "Y") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_small, p_optimal, p_large, ncol = 3)
```

## Bandwidth vs. Cluster Count Analysis

```{r exp3_bandwidth_analysis, fig.height=5}
# Test range of bandwidth values
bandwidth_values <- seq(0.2, 2.0, by = 0.5)
n_clusters_found <- numeric(length(bandwidth_values))

for (i in seq_along(bandwidth_values)) {
  result_temp <- mean_shift_clustering(
    dense_data$data,
    bandwidth = bandwidth_values[i],
    kernel = "gaussian"
  )
  n_clusters_found[i] <- result_temp$n_clusters
}

# Plot
ggplot(data.frame(bandwidth = bandwidth_values, clusters = n_clusters_found)) +
  geom_line(aes(x = bandwidth, y = clusters), size = 1, color = "blue") +
  geom_point(aes(x = bandwidth, y = clusters), size = 2) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 2.5, y = 3.5, label = "True clusters = 3", color = "red") +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "darkgreen", size = 1) +
  annotate("text", x = 1.2, y = max(n_clusters_found) - 1,
           label = "Optimal h = 1.0", color = "darkgreen") +
  labs(title = "Bandwidth vs. Number of Clusters Found",
       x = "Bandwidth (h)",
       y = "Number of Clusters") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

## Critical Insights

- **Too small bandwidth → over-segmentation**: Many tiny clusters instead of true 3 clusters
- **Too large bandwidth → under-segmentation**: All points merge into 1 cluster
- **Bandwidth is the most critical parameter**: More important than kernel choice
- **Automatic selection methods**: Scott's and Silverman's rules provide good starting points

# Bandwidth Selection Strategies

```{r bandwidth_strategies}
# Compare automatic bandwidth selection methods
bandwidth_scott <- scotts_rule(medium_data$data)
bandwidth_silverman <- silvermans_rule(medium_data$data)

cat(sprintf("Scott's Rule: h = %.4f\n", bandwidth_scott))
cat(sprintf("Silverman's Rule: h = %.4f\n", bandwidth_silverman))
cat(sprintf("Manually chosen optimal: h = 1.0\n"))

# Test both automatic methods
result_scott <- mean_shift_clustering(
  dense_data$data,
  bandwidth = bandwidth_scott,
  kernel = "gaussian"
)

result_silverman <- mean_shift_clustering(
  dense_data$data,
  bandwidth = bandwidth_silverman,
  kernel = "gaussian"
)

bandwidth_strategies <- rbind(
  cbind(Method = "Scott's Rule", h = round(bandwidth_scott, 3),
        evaluate_clustering(medium_data$data, result_scott, medium_data$labels)),
  cbind(Method = "Silverman's Rule", h = round(bandwidth_silverman, 3),
        evaluate_clustering(medium_data$data, result_silverman, medium_data$labels)),
  cbind(Method = "Manual Optimal", h = 1.0,
        evaluate_clustering(medium_data$data, result_optimal, medium_data$labels))
)

knitr::kable(bandwidth_strategies, caption = "Bandwidth Selection Strategies Comparison")
```

# Experiment 4: Scalability Analysis

**Objective**: Analyze how runtime and computational cost scale with dataset size to understand practical limitations.

## Setup

We test the algorithm on datasets ranging from 100 to 10,000 points with 3 well-separated clusters. This allows us to observe the O(n²) computational complexity in practice.

```{r exp4_setup}
# Dataset sizes to test
dataset_sizes <- c(100, 250, 500, 750, 1000, 2000, 3000)

# Use consistent cluster configuration
cluster_centers <- matrix(c(0, 0, 5, 5, 10, 0), ncol = 2, byrow = TRUE)
cluster_std <- 0.5
bandwidth_fixed <- 0.8
kernel_fixed <- "gaussian"

# Initialize results storage
scalability_results <- data.frame(
  n_samples = integer(),
  runtime_secs = numeric(),
  mean_iterations = numeric(),
  n_clusters = integer()
)

# Run experiments
set.seed(42)
for (n in dataset_sizes) {
  cat(sprintf("Testing n = %d...\n", n))

  # Generate data
  data_blob <- generate_blobs(
    n_samples = n,
    centers = cluster_centers,
    cluster_std = cluster_std
  )

  # Run mean shift
  result <- mean_shift_clustering(
    data_blob$data,
    bandwidth = bandwidth_fixed,
    kernel = kernel_fixed
  )

  # Store results
  scalability_results <- rbind(
    scalability_results,
    data.frame(
      n_samples = n,
      runtime_secs = result$computation_time,
      mean_iterations = result$mean_iterations,
      n_clusters = result$n_clusters
    )
  )
}
```

## Results Table

```{r exp4_results_table}
knitr::kable(scalability_results,
             caption = "Experiment 4: Runtime Scalability with Dataset Size",
             digits = 3)
```

## Visualization: Runtime vs Dataset Size

```{r exp4_viz_runtime, fig.height=5}
# Plot runtime vs dataset size
ggplot(scalability_results, aes(x = n_samples, y = runtime_secs)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Runtime vs Dataset Size",
    subtitle = "Mean Shift Clustering with Gaussian kernel (h = 0.8)",
    x = "Number of Data Points (n)",
    y = "Runtime (seconds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

## Complexity Analysis: Quadratic Scaling

```{r exp4_viz_complexity, fig.height=5}
# Add theoretical O(n²) curve for comparison
scalability_results$n_squared_normalized <- (scalability_results$n_samples^2) /
  (scalability_results$n_samples[1]^2) * scalability_results$runtime_secs[1]

ggplot(scalability_results) +
  geom_line(aes(x = n_samples, y = runtime_secs, color = "Observed"),
            size = 1.2) +
  geom_point(aes(x = n_samples, y = runtime_secs, color = "Observed"),
             size = 3) +
  geom_line(aes(x = n_samples, y = n_squared_normalized, color = "O(n²) theoretical"),
            size = 1, linetype = "dashed") +
  scale_color_manual(
    values = c("Observed" = "blue", "O(n²) theoretical" = "red"),
    name = "Scaling"
  ) +
  labs(
    title = "Observed Runtime vs Theoretical O(n²) Complexity",
    x = "Number of Data Points (n)",
    y = "Runtime (seconds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.position = "bottom"
  )
```

## Mean Iterations vs Dataset Size

```{r exp4_viz_iterations, fig.height=5}
ggplot(scalability_results, aes(x = n_samples, y = mean_iterations)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(color = "green", size = 3) +
  labs(
    title = "Mean Iterations per Point vs Dataset Size",
    subtitle = "Number of iterations until convergence",
    x = "Number of Data Points (n)",
    y = "Mean Iterations"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

## Key Insights

- **Quadratic scaling confirmed**: Runtime grows approximately as O(n²), matching theoretical complexity
- **Practical limit**: For n > 10,000 points, runtime becomes impractical without optimizations
- **Iterations remain stable**: Mean iterations per point stays relatively constant regardless of dataset size
- **Bottleneck**: The pairwise distance computation in each iteration dominates the runtime
- **Recommendations for large datasets**:
  - Use spatial indexing structures (KD-trees, Ball trees)
  - Consider approximate methods or sampling techniques
  - Implement parallelization for independent point updates

# Summary of Experimental Findings

## Key Takeaways

1. **Data Characteristics**:
   - Works well on well-separated clusters
   - Struggles with heavy overlap
   - Density-based approach can handle moderate overlap effectively

2. **Kernel Choice**:
   - Flat/Uniform kernel: fastest, good results
   - Gaussian: slightly better quality, slower
   - Choice matters less than bandwidth selection

3. **Bandwidth Selection** (MOST CRITICAL):
   - Most important parameter affecting results
   - Use adaptive methods (Scott's/Silverman's rule) as starting point
   - Fine-tune based on domain knowledge or cross-validation
   - Validate with silhouette score

4. **Scalability and Performance**:
   - Runtime scales quadratically with dataset size: O(n²)
   - Practical limit around 10,000 points without optimization
   - Mean iterations remain stable regardless of dataset size
   - Pairwise distance computation is the main bottleneck

## Practical Recommendations

- Start with automatic bandwidth selection (Scott's or Silverman's rule)
- Visualize results and adjust if needed
- For well-separated clusters, use slightly larger bandwidth
- For overlapping data, use smaller bandwidth but expect more clusters
- **For large datasets (n > 10,000)**:
  - Consider spatial indexing (KD-trees, Ball trees) to reduce complexity
  - Use sampling or approximate methods
  - Implement parallel computation for point updates
  - Consider alternative algorithms (DBSCAN, Hierarchical clustering) for very large datasets

# Session Info

```{r session_info}
sessionInfo()
```
